An Improved Approximate-Bayesian Model-choice Method for Estimating Shared Evolutionary History

### 2014-02-23 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
More editing of DPP ms.
Incorporating Hickerson's and our response papers into the intro.

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
This causes divergence models with more divergence-time parameters to integrate
over a \emph{much} greater parameter space with low likelihood yet high prior
density, resulting in small marginal likelihoods relative to models with fewer
divergence-time parameters \citep{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this can cause support for models with
fewer divergence events \citep{Oaks2012}.
Furthermore, \msb uses a discrete uniform prior over the number of divergence
events $1,2,\ldots,\npairs{}$.
Because there are many more possible assignments of populations pairs to
intermediate numbers of divergence events, this imposes a prior on divergence
models that puts most of the prior mass on models with either very few or very
many divergence-time parameters (see Figure 5 of \citet{Oaks2012}; for brevity
I will refer to this prior as ``U-shaped'').
Given that models with many divergence events can have small marginal
likelihoods due to the uniform priors on divergence-time parameters, the
U-shaped prior will effectively create a strong prior preference for models
with very few divergence events.
Recently, \citet{Oaks2012} found via simulation that \msb will often strongly
support biogeographically interesting models with a small number of divergence
events shared among the taxa, even when divergences were random over broad
timescales.
They suggested this behavior was due to the combination of uniform priors on
parameters and the U-shaped prior on divergence models.

@@

Here, I modify the \msb model to test whether alternative parameterizations and
priors will improve the behavior of the approximate-Bayesian model-choice
approach.
I implement a Dirichlet-process prior (DPP) over all possible models of
divergence, and use gamma and beta probability distributions in place of
uniform priors on many of the model's parameters.
Using simulations, I show that the new implementation has improved robustness,
accuracy, and power compared to the original model.
Also, my results confirm that the improved performance of the new model
is due to a combination of
(1) more flexible priors on divergence-time and demographic parameters that
avoid placing high prior density in improbable regions of parameter space, and
(2) a diffuse Dirichlet-process prior that does not strongly disfavor
divergence models with intermediate numbers of divergence events.
% Also, our results confirm that both the uniform priors on demographic and
% divergence-time parameters and the U-shaped prior on divergence models
% contribute to the tendency of \msb.
After reanalyzing sequence data from 22 pairs of taxa from the Philippines
\citep{Oaks2012} under the new model, I find a large amount of posterior
uncertainty in the number of divergence events shared among the taxa; a result
in contrast with the original \msb model and congruent with
intuition given the richness of the model and the relatively small amount of
information in the data.

>>>>

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
This causes divergence models with more divergence-time parameters to integrate
over a \emph{much} greater parameter space with low likelihood yet high prior
density, resulting in small marginal likelihoods relative to models with fewer
divergence-time parameters \citep{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this can cause support for models with
fewer divergence events \citep{Oaks2012,Oaks2014reply}.
Furthermore, \msb uses a discrete uniform prior over the number of divergence
events $1,2,\ldots,\npairs{}$.
Because there are many more possible assignments of populations pairs to
intermediate numbers of divergence events, this imposes a prior on divergence
models that puts most of the prior mass on models with either very few or very
many divergence-time parameters (see Figure 5 of \citet{Oaks2012}; for brevity
I will refer to this prior as ``U-shaped'').
Given that models with many divergence events can have small marginal
likelihoods due to the uniform priors on divergence-time parameters, the
U-shaped prior will effectively create a strong prior preference for models
with very few divergence events.

Recently, \citet{Oaks2012} and \citet{Oaks2014reply} found via simulation that
\msb will often strongly support biogeographically interesting models with a
small number of divergence events shared among taxa, even when divergences were
random over broad timescales.
They suggested this behavior was due to the combination of uniform priors on
parameters causing small marginal likelihoods of richer models and the
U-shaped prior on divergence models.
\citet{Hickerson2013} suggested the problem is caused by sampling error, and
proposed as a solution an approximate-Bayesian model averaging approach that
samples over empirically informed uniform priors.
However, \citet{Oaks2014reply} evaluated the approach proposed by
\citet{Hickerson2013} using simulations and found that it did not mitigate the
method's propensity to incorrectly infer clustered divergences, and often
preferred priors that excluded the true values of the model's parameters.
Here, I describe a new method that successfully mitigates spurious inference
of biogeographically interesting scenarios while avoiding negative side effects
of empirically informed uniform priors.

@@

In this study, I introduce a new method, implemented in the software
\dppmsbayes, that extends the model of \msb.
I use this method to test whether alternative parameterizations and priors
improve the behavior of the approximate-Bayesian model-choice approach to
estimating shared divergence events.
The new approach uses a Dirichlet-process prior (DPP) over all possible models
of divergence, and gamma and beta probability distributions in place of uniform
priors on many of the model's parameters.
Using simulations, I show that the new implementation has improved robustness,
accuracy, and power compared to the original model.
Also, my results confirm that the improved performance of the new model
is due to a combination of
(1) more flexible priors on divergence-time and demographic parameters that
avoid placing high prior density in improbable regions of parameter space, and
(2) a diffuse Dirichlet-process prior that does not strongly disfavor
divergence models with intermediate numbers of divergence events.
% Also, our results confirm that both the uniform priors on demographic and
% divergence-time parameters and the U-shaped prior on divergence models
% contribute to the tendency of \msb.
After reanalyzing sequence data from 22 pairs of taxa from the Philippines
\citep{Oaks2012} under the new model, I find a large amount of posterior
uncertainty in the number of divergence events shared among the taxa; a result
in contrast with the original \msb model and congruent with
intuition given the richness of the model and the relatively small amount of
information in the data.

### 2014-02-23 manuscripts/dpp-msbayes/dpp-msbayes.tex
More editing of DPP ms.
Incorporating Hickerson's and our response papers into the intro.

\begin{abstract}
    To understand the processes that generate biodiversity, it is
    important to account for large-scale processes that affect the evolutionary
    history of groups of co-distributed populations of organisms.
    Such events predict temporally clustered divergences times, a pattern
    that can be estimated using genetic data from co-distributed species.
    I introduce a new approximate-Bayesian model for comparative
    phylogeographical model-choice that estimates the temporal distribution of
    divergences across taxa from multi-locus DNA sequence data.
    The model is an extension of that implemented in \msb.
    By reparameterizing the model, introducing more flexible priors on
    demographic and divergence-time parameters, and implementing a
    non-parametric Dirichlet-process prior over divergence models, I improved
    the robustness, accuracy, and power of the method for estimating shared
    evolutionary history across taxa.
    The results demonstrate the improved performance of the new method is due
    to (1) more appropriate priors on divergence-time and demographic
    parameters that avoid prohibitively small marginal likelihoods for models
    with more divergence events,
    and (2) the Dirichlet-process providing a flexible prior on divergence
    histories that does not strongly disfavor models with intermediate numbers
    of divergence events.
    % The results demonstrate that the tendency of the \msb model to spuriously
    % support models of clustered divergences is caused by a combination of (1)
    % uniform priors on nuisance parameters hindering the marginal likelihoods of
    % models with more divergence-time parameters, and (2) a prior on divergence
    % models that disfavors models with intermediate numbers of divergence-time
    % parameters.
    The new method yields more robust estimates of posterior uncertainty, and
    thus greatly reduces the tendency of the model to incorrectly estimate
    biogeographically interesting models with strong support.

>>>>

\begin{abstract}
    To understand the processes that generate biodiversity, it is
    important to account for large-scale processes that affect the evolutionary
    history of groups of co-distributed populations of organisms.
    Such events predict temporally clustered divergences times, a pattern
    that can be estimated using genetic data from co-distributed species.
    I introduce a new approximate-Bayesian method for comparative
    phylogeographical model-choice that estimates the temporal distribution of
    divergences across taxa from multi-locus DNA sequence data.
    The model is an extension of that implemented in \msb.
    By reparameterizing the model, introducing more flexible priors on
    demographic and divergence-time parameters, and implementing a
    non-parametric Dirichlet-process prior over divergence models, I improved
    the robustness, accuracy, and power of the method for estimating shared
    evolutionary history across taxa.
    The results demonstrate the improved performance of the new method is due
    to (1) more appropriate priors on divergence-time and demographic
    parameters that avoid prohibitively small marginal likelihoods for models
    with more divergence events,
    and (2) the Dirichlet-process providing a flexible prior on divergence
    histories that does not strongly disfavor models with intermediate numbers
    of divergence events.
    % The results demonstrate that the tendency of the \msb model to spuriously
    % support models of clustered divergences is caused by a combination of (1)
    % uniform priors on nuisance parameters hindering the marginal likelihoods of
    % models with more divergence-time parameters, and (2) a prior on divergence
    % models that disfavors models with intermediate numbers of divergence-time
    % parameters.
    The new method yields more robust estimates of posterior uncertainty, and
    thus greatly reduces the tendency of the model to incorrectly estimate
    biogeographically interesting models with strong support.
